---
title: 'Foundations for statistical inference - Confidence intervals'
author: "Kayleah Griffen"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

If you have access to data on an entire population, say the opinion of every adult in the United States on whether or not they think climate change is affecting their local community, it's straightforward to answer questions like, "What percent of US adults think climate change is affecting their local community?". Similarly, if you had demographic information on the population you could examine how, if at all, this opinion varies among young and old adults and adults with different leanings. If you have access to only a sample of the population, as is often the case, the task becomes more complicated. What is your best guess for this proportion if you only have data from a small sample of adults? This type of situation requires that you use your sample to make inference on what your population looks like.

<div id="boxedtext">
**Setting a seed:** You will take random samples and build sampling distributions in this lab, which means you should set a seed on top of your lab. If this concept is new to you, review the lab on probability.
</div>

```{r}
set.seed(9041996)
```

## Getting Started

### Load packages

In this lab, we will explore and visualize the data using the **tidyverse** suite of packages, and perform statistical inference using **infer**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(infer)
```

### The data

A 2019 Pew Research report states the following:

To keep our computation simple, we will assume a total population size of 100,000 (even though that's smaller than the population size of all US adults).

> Roughly six-in-ten U.S. adults (62%) say climate change is currently affecting their local community either a great deal or some, according to a new Pew Research Center survey.
>
>**Source:** [Most Americans say climate change impacts their community, but effects vary by region](https://www.pewresearch.org/fact-tank/2019/12/02/most-americans-say-climate-change-impacts-their-community-but-effects-vary-by-region/)

In this lab, you will assume this 62% is a true population proportion and learn about how sample proportions can vary from sample to sample by taking smaller samples from the population. We will first create our population assuming a population size of 100,000. This means 62,000 (62%) of the adult population think climate change impacts their community, and the remaining 38,000 does not think so.

```{r}
us_adults <- tibble(
  climate_change_affects = c(rep("Yes", 62000), rep("No", 38000))
)
```

The name of the data frame is `us_adults` and the name of the variable that contains responses to the question *"Do you think climate change is affecting your local community?"* is `climate_change_affects`.

We can quickly visualize the distribution of these responses using a bar plot.

```{r bar-plot-pop, fig.height=2.5, fig.width=10}
ggplot(us_adults, aes(x = climate_change_affects)) +
  geom_bar() +
  labs(
    x = "", y = "",
    title = "Do you think climate change is affecting your local community?"
  ) +
  coord_flip() 
```

We can also obtain summary statistics to confirm we constructed the data frame correctly.

```{r summ-stat-pop, results = TRUE}
us_adults %>%
  count(climate_change_affects) %>%
  mutate(p = n /sum(n))
```

In this lab, you'll start with a simple random sample of size 60 from the population.

```{r sample}
n <- 60
samp <- us_adults %>%
  sample_n(size = n)
```

1.  What percent of the adults in your sample think climate change affects their 
    local community? **Hint:** Just like we did with the population, we can calculate 
    the proportion of those **in this sample** who think climate change affects their 
    local community.

```{r}
samp %>%
  count(climate_change_affects) %>%
  mutate(p = n /sum(n))
```

In my sample 63% of adults think that climate change affects their local community.

2.  Would you expect another student's sample proportion to be identical to yours? 
    Would you expect it to be similar? Why or why not?
    
I would expect that another students sample proportion be similar to mine, but not identical. Depending on our seeds and what calculations involving random numbers were done prior to running - the sample would most likely be different. I would expect that they are similar, the sample size is relatively small but still the sample is random and it should semi-closely resemble the population.

## Confidence intervals

Return for a moment to the question that first motivated this lab: based on this sample, what can you infer about the population? With just one sample, the best estimate of the proportion of US adults who think climate change affects their local community would be the sample proportion, usually denoted as $\hat{p}$ (here we are calling it `p_hat`). That serves as a good **point estimate**, but it would be useful to also communicate how uncertain you are of that estimate. This uncertainty can be quantified using a **confidence interval**.

One way of calculating a confidence interval for a population proportion is based on the Central Limit Theorem, as $\hat{p} \pm z^\star SE_{\hat{p}}$ is, or more precisely, as \[ \hat{p} \pm z^\star \sqrt{ \frac{\hat{p} (1-\hat{p})}{n} } \]

Another way is using simulation, or to be more specific, using **bootstrapping**. The term **bootstrapping** comes from the phrase "pulling oneself up by one's bootstraps", which is a metaphor for accomplishing an impossible task without any outside help. In this case the impossible task is estimating a population parameter (the unknown population proportion), and we'll accomplish it using data from only the given sample. Note that this notion of saying something about a population parameter using only information from an observed sample is the crux of statistical inference, it is not limited to bootstrapping. 

In essence, bootstrapping assumes that there are more of observations in the populations like the ones in the observed sample. So we "reconstruct" the population by resampling from our sample, with replacement. The bootstrapping scheme is as follows:

- **Step 1.** Take a bootstrap sample - a random sample taken **with replacement** from the original sample, of the same size as the original sample.
- **Step 2.** Calculate the bootstrap statistic - a statistic such as mean, median, proportion, slope, etc. computed on the bootstrap samples.
- **Step 3.** Repeat steps (1) and (2) many times to create a bootstrap distribution - a distribution of bootstrap statistics.
- **Step 4.** Calculate the bounds of the XX% confidence interval as the middle XX% j knof the bootstrap distribution.

Instead of coding up each of these steps, we will construct confidence intervals using the **infer** package.

Below is an overview of the functions we will use to construct this confidence interval:

Function    | Purpose
----------- | -------
`specify`   | Identify your variable of interest
`generate`  | The number of samples you want to generate
`calculate` | The sample statistic you want to do inference with, or you can also think of this as the population parameter you want to do inference for
`get_ci`    | Find the confidence interval

This code will find the 95 percent confidence interval for proportion of US adults who think climate change affects their local community.

```{r confidence interval infer}
samp %>%
  specify(response = climate_change_affects, success = "Yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = 0.95)
```

- In `specify` we specify the `response` variable and the level of that variable we are calling a `success`.
- In `generate` we provide the number of resamples we want from the population in the `reps` argument (this should be a reasonably large number) as well as the type of resampling we want to do, which is `"bootstrap"` in the case of constructing a confidence interval.
- Then, we `calculate` the sample statistic of interest for each of these resamples, which is `prop`ortion.

Feel free to test out the rest of the arguments for these functions, since these commands will be used together to calculate confidence intervals and solve inference problems for the rest of the semester. But we will also walk you through more examples in future chapters.

To recap: even though we don't know what the full population looks like, we're 95% confident that the true proportion of US adults who think climate change affects their local community is between the two bounds reported as result of this pipeline.

## Confidence levels

3.  In the interpretation above, we used the phrase "95% confident". What does "95% confidence" mean?

According to the OS textbook, a confidence interval represents a range of values where it is likely that the population parameter will be. To be able to construct the 95% confidence interval the point estimate has to qualify for central limit theorem. In a normal distribution 95% of the data is within 1.96 standard deviations of the mean. Using this knowledge the confidence interval can be calculated using the point estimate +/- 1.96*SE. 

Basically in summary, the 95% confidence interval is based on taking the point estimate and going +/- 1.96 standard deviations because as long as the data follows a normal distribution 95% of the data points should be +/- 1.96 from the mean. 

In this case, you have the rare luxury of knowing the true population proportion (62%) since you have data on the entire population. 

4.  Does your confidence interval capture the true population proportion of US adults 
    who think climate change affects their local community? If you are working on this 
    lab in a classroom, does your neighbor's interval capture this value? 
    
My confidence interval captures the true proportion of US adults who think climate change effects their community. My lower CI was 51.6% and upper CI was 75%, 62% (the true population proportion) falls inbetween these.

5.  Each student should have gotten a slightly different confidence interval. What 
    proportion of those intervals would you expect to capture the true population 
    mean? Why?

Each students CI would be slightly different because it is based on their sample data - which as previously noted would be different due to different seeds. I would expect almost all of them (95%) to capture the true population mean. 

In the next part of the lab, you will collect many samples to learn more about how sample proportions and confidence intervals constructed based on those samples vary from one sample to another.

-   Obtain a random sample.
-   Calculate the sample proportion, and use these to calculate and store the lower and upper bounds of the confidence intervals.
-   Repeat these steps 50 times.

Doing this would require learning programming concepts like iteration so that you can automate repeating running the code you've developed so far many times to obtain many (50) confidence intervals. In order to keep the programming simpler, we are providing the interactive app below that basically does this for you and created a plot similar to Figure 5.6 on [OpenIntro Statistics, 4th Edition (page 182)](https://www.openintro.org/os).

```{r shiny, echo=FALSE, eval=FALSE, results = TRUE}
# This R chunk will only run in interactive mode
library(shiny)
store_ci <- function(i, n, reps, conf_level, success) {
  us_adults %>%
    sample_n(size = n) %>%
    specify(response = climate_change_affects, success = success) %>%
    generate(reps, type = "bootstrap") %>%
    calculate(stat = "prop") %>%
    get_ci(level = conf_level) %>%
    rename(
      x_lower = names(.)[1],
      x_upper = names(.)[2]
    )
}
library(shiny)
shinyApp(
  ui <- fluidPage(
    h4("Confidence intervals for the proportion of US adults who think 
     climate change"),

    h4(selectInput("success", "",
      choices = c(
        "is affecting their local community" = "Yes",
        "is not affecting their local community" = "No"
      ),
      selected = "Yes", width = "50%"
    )),

    # Sidebar with a slider input for number of bins
    sidebarLayout(
      sidebarPanel(
        numericInput("n_samp",
          "Sample size for a single sample from the population:",
          min = 1,
          max = 1000,
          value = 60
        ),

        hr(),

        numericInput("n_rep",
          "Number of resamples for each bootstrap confidence interval:",
          min = 1,
          max = 15000,
          value = 1000
        ),

        numericInput("conf_level",
          "Confidence level",
          min = 0.01,
          max = 0.99,
          value = 0.95,
          step = 0.05
        ),

        hr(),

        radioButtons("n_ci",
          "Number of confidence intervals:",
          choices = c(10, 25, 50, 100),
          selected = 50, inline = TRUE
        ),

        actionButton("go", "Go")
      ),

      # Show a plot of the generated distribution
      mainPanel(
        plotOutput("ci_plot")
      )
    )
  ),

  server <- function(input, output) {

    # set true p
    p <- reactive(ifelse(input$success == "Yes", 0.62, 0.38))

    # create df_ci when go button is pushed
    df_ci <- eventReactive(input$go, {
      map_dfr(1:input$n_ci, store_ci,
        n = input$n_samp,
        reps = input$n_rep, conf_level = input$conf_level,
        success = input$success
      ) %>%
        mutate(
          y_lower = 1:input$n_ci,
          y_upper = 1:input$n_ci,
          capture_p = ifelse(x_lower < p() & x_upper > p(), "Yes", "No")
        )
    })

    # plot df_ci
    output$ci_plot <- renderPlot({
      ggplot(df_ci()) +
        geom_segment(aes(x = x_lower, y = y_lower, xend = x_upper, yend = y_upper, color = capture_p)) +
        geom_point(aes(x = x_lower, y = y_lower, color = capture_p)) +
        geom_point(aes(x = x_upper, y = y_upper, color = capture_p)) +
        geom_vline(xintercept = p(), color = "darkgray") +
        labs(
          y = "", x = "Bounds of the confidence interval",
          color = "Does the interval capture the true population proportion?"
        ) +
        theme(legend.position = "bottom")
    })
  },
  options = list(height = 700)
)
```


6.  Given a sample size of 60, 1000 bootstrap samples for each interval, and 50 
    confidence intervals constructed (the default values for the above app), what 
    proportion of your confidence intervals include the true population proportion? 
    Is this proportion exactly equal to the confidence level? If not, explain why. 
    Make sure to include your plot in your answer.

To get the plot for this graph you have to run the Shiny app with the default settings. Given 50 confidence intervals, 47/50 include the true population proportion. That would be 94% of the CIs include the true population interval - this is basically 95% so its essentially equal to the confidence interval. 

* * *

## More Practice

7.  Choose a different confidence level than 95%. Would you expect a confidence 
    interval at this level to me wider or narrower than the confidence interval 
    you calculated at the 95% confidence level? Explain your reasoning.
    
I chose a 50% CI. This confidence interval is narrower, meaning that my range of values between the CIs will be smaller than what it was for 95% (which had a bigger range). The trade off for this narrowing in is that the proportion of my confidence intervals that include the true population proportion is less than the 95% - closer to 50%.

8.  Using code from the **infer** package and data from the one sample you have 
    (`samp`), find a confidence interval for the proportion 
    of US Adults who think climate change is affecting their local community with a 
    confidence level of your choosing (other than 95%) and interpret it.
    
```{r}
samp %>%
  specify(response = climate_change_affects, success = "Yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = 0.50)
```

The 50% confidence interval lower bound is 0.596 and upper bound is 0.667. Compared to the 95% confidence interval lower bound of 0.516 and upper bound of 0.75 - the 50% confidence interval is narrower - but the tradeoff is that the true population proportion is less likely to be in the 50% confidence interval. In my case however, my confidence interval of 50% did capture the population mean.

9.  Using the app, calculate 50 confidence intervals at the confidence level you chose 
    in the previous question, and plot all intervals on one plot, and calculate 
    the proportion of intervals that include the true population proportion. 
    How does this percentage compare to the confidence level selected for the 
    intervals?
    
In my plot 26/50 of the intervals include the true population proportion. This percentage (52%) is about equal to the confidence interval I chose, 50%.

    
10.  Lastly, try one more (different) confidence level. First, state how you expect the
    width of this interval to compare to previous ones you calculated. Then, 
    calculate the bounds of the interval using the **infer** package and data 
    from `samp` and interpret it. Finally, use the app to generate many intervals 
    and calculate the proportion of intervals that are capture the true population 
    proportion.
    
```{r}
samp %>%
  specify(response = climate_change_affects, success = "Yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = 0.75)
```

For this test I tried the 75% confidence interval. I expect that the 75% confidence interval will be wider than the 50% and narrower than the 95%. My lower CI was 0.567 and upper was 0.7. - as expected narrower than the 95% CI but wider than the 50%. I expect about 38/50 of the intervals in my plot to include the true population proportion. When I ran the app I got 39/50 of my intervals (78%) captured the true population proportion.

11.  Using the app, experiment with different sample sizes and comment on how the 
    widths of intervals change as sample size changes (increases and decreases).
    
With a greater sample sizes the width of the interval decreases because the SE decreases.  
    
12.  Finally, given a sample size (say, 60), how does the width of the interval change 
    as you increase the number of bootstrap samples. **Hint:** Does changing the 
    number of bootstap samples affect the standard error?
    
As you increase the number of bootstrap samples the SE does not change significantly (changing the sample size would decrease the SE). I'd expect that the SE from the samples is relatively similar so changing the amount of bootstrap samples you take would not effect the SE. Therefor the CI width would be the same as you increase bootstrap samples. 

* * *